An exposition on the propriety of restricted Boltzmann machines

### 2016-08-11 writing/draft.Rmd
slides complete, but check densities
update document to JSM content

---
title: "An exposition on the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Stephen Vardeman, Daniel Nordman"
date: "June, 2016"
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
---

@@

@@

#Characterizations
##Tiny example
To illustrate the ideas of graph near-degeneracy, instability, and uninterpretability in a RBM, we will consider a the smallest possible toy example that consists of one visible node $v_1$ and one hidden node $h_1$, both binary. A schematic of this model can be found in figure \ref{fig:toymodel}.

@@

#Bayesian model fitting
To avoid model impropriety, we want to avoid parts of the parameter space that lead to near-degeneracy}, instability, and uninterpretability. Based on the evidence in section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta$ by specifying priors that place low probability on high values of $\boldsymbol \theta$.

@@

1. *Trick prior.* Cancel out the normalizing term, resulting full conditionals are normally distributed. 
    \begin{align*}
    \pi(\boldsymbol \theta) \propto \gamma(\boldsymbol \theta)^n \exp\left(-\frac{1}{2C_{1}}\boldsymbol \theta_{main}'\boldsymbol \theta_{main} -\frac{1}{2C_{2}}\boldsymbol \theta_{interaction}'\boldsymbol \theta_{interaction}\right), \vspace{-.75cm}
    \end{align*}

@@

The three fitting methods are ordered by computational feasibility in a real data situation with the trick prior being the most possible due to conjugacy and the marginalized likelihood method the least feasible due to the marginalization and need for an adaptive Metropolis Hastings step. The trick prior, however, does require extensive tuning to choose $C_1$ and $C_2$, making it less appealing.

Figure \ref{fig:fitting_plot} shows the posterior probability of each possible binary image with four pixels after fitting the model in the three ways detailed in this section. The black lines show the true probabilities of each image based on the parameters used to simulate the images. From these posterior predictive checks, it is evident that the marginalized likelihood method shows the closest fit to the actual data. However, this model requires a marginalization step, which is infeasible for a model of any real size.

```{r fitting}
#data and params ------------------------
H <- 4
V <- 4

#marginalized likelihood
load("data/fitted_models_adaptive_mh_trunc_distn_marginal_1.Rdata")
marginal_bad_1 <- models_bad
marginal_good_1 <- models_good

#load Jing's prior
load("data/fitted_models_distn_0.26.Rdata")
jing_bad <- models_bad
jing_good <- models_good

#truncated normal
load("data/fitted_models_adaptive_mh_trunc_distn_shrunk_jing_match.Rdata")
shrunk_jing_bad <- models_bad
shrunk_jing_good <- models_good

#rm unneccesary data
rm(models_bad)
rm(models_degen)
rm(models_good)

load("data/sample.params.Rdata")
params_degen <- list(main_hidden = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
                     main_visible = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
                     interaction = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))

load("data/sample_images.Rdata")

#computing actual distributions ---------------------
distn_good <- visible_distn(params = params_good)
distn_degen <- visible_distn(params = params_degen)

reshape_sample_distn <- function(model) {
  sample_distn <- model$distn
  dim(sample_distn) <- c(dim(sample_distn)[1]*dim(sample_distn)[2], dim(sample_distn)[3])
  sample_distn %>% data.frame() -> sample_distn
  names(sample_distn) <- names(distn_degen)

@@

marginal_sample_good_1 <- reshape_sample_distn(marginal_good_1)
jing_sample_good <- reshape_sample_distn(jing_good)
shrunk_jing_sample_good <- reshape_sample_distn(shrunk_jing_good) %>% filter(iter > 500) %>% mutate(iter = 1:n())

#plots -------------------
marginal_sample_good_1 %>% rename(marginal = prob) %>%
  left_join(jing_sample_good %>% ungroup() %>% select(-image_id) %>% rename(trick = prob)) %>%
  left_join(shrunk_jing_sample_good %>% ungroup() %>% select(-image_id) %>% rename(trunc = prob)) %>%
  ungroup() %>%
  mutate(method = "good", image_id = paste0("image_", image_id)) %>% 
  #select_(.dots = paste0("-v", 1:V)) %>%
  gather(prior, prob, trick, marginal, trunc) %>%
  spread(image_id, prob) -> all_statistics

all_statistics %>%
  gather(statistic, value, -iter, -method, -prior, -starts_with("v")) %>%
  filter(grepl("image", statistic) & !is.na(value)) %>%
  separate(statistic, into = c("junk", "statistic")) %>%
  mutate(statistic = factor(statistic, labels = paste("Image", 1:length(unique(statistic))))) %>%
  left_join(distn_good %>% select(-image_id)) %>% 
  ggplot() +
  geom_line(aes(iter/100, value, colour = prior)) +
  geom_abline(aes(slope = 0, intercept = prob)) +
  facet_wrap(~statistic) +
  ylim(c(0,.5)) +
  xlab("MCMC Iteration (in 100s)") + ylab("Posterior Probability") +
  theme(legend.position = "bottom") +
  scale_colour_discrete("Method", labels = c("Marginalized likelihood", "Trick prior", "Truncated Normal prior")) -> p

```{r fitting_plot, fig.cap="\\label{fig:fitting_plot}Posterior probability of each possible 4-pixel image using three Bayesian fitting techniques, trick prior, truncated Normal prior, and marginalized likelihood. The black lines show the true probabilities of each image based on the parameters used to simulate the images. The marginalized likelihood method has the best fit for the data, however is also the most computationally intensive and least feasible with a real dataset.", fig.height=6}
p
```

@@

**TODO: Add proof that likelihood goes to infinity.**  
**TODO: Add comparison to contrastive divergence**

@@

>>>>

---
title: "An exposition on the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Stephen Vardeman, Daniel Nordman"
date: "August, 2016"
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
---

@@

Thus, if a discrete exponential family model is unstable, then it is also degenerate.

@@

#Simulations

We explore the relationship between magnitudes of $\boldsymbol \theta$ and the three measures of model impropriety, near-degeneracy, instability, and uninterpretability, for varying sized RBM models through simulated data.

##Tiny example
To illustrate the ideas of graph near-degeneracy, instability, and uninterpretability in a RBM, we will consider a the smallest possible toy example that consists of one visible node $v_1$ and one hidden node $h_1$, both binary. A schematic of this model can be found in figure \ref{fig:toymodel}.

@@

#Bayesian model fitting
To avoid model impropriety, we want to avoid parts of the parameter space that lead to near-degeneracy, instability, and uninterpretability. Based on the evidence in section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta$ by specifying priors that place low probability on high values of $\boldsymbol \theta$, specifically focusing on shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.

@@

1. *Trick prior.* Cancel out normalizing term, resulting full conditionals normally distributed. 
    \begin{align*}
    \pi(\boldsymbol \theta) \propto \gamma(\boldsymbol \theta)^n \exp\left(-\frac{1}{2C_{1}}\boldsymbol \theta_{main}'\boldsymbol \theta_{main} -\frac{1}{2C_{2}}\boldsymbol \theta_{interaction}'\boldsymbol \theta_{interaction}\right), \vspace{-.75cm}
    \end{align*}

@@

The three fitting methods are ordered by computational feasibility in a real data situation with the trick prior being the most possible due to conjugacy and the marginalized likelihood method the least feasible due to the marginalization and need for an adaptive Metropolis Hastings step. All three priors do require choosing the values of hyperparameters. In each case, we have chosed the values based on a rule of thumb that maintains the property of shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$. Additionally, the trick prior requires additional tuning to choose $C_1$ and $C_2$, making it less appealing. The values used for the hyperparameters are presented in table \ref{tab:hyperparam}.

\begin{table}[ht]
\centering
\begin{tabular}{|l|c|c|}
\hline 
Method & Hyperparameter & Value \\ 
\hline \hline
\multirow{2}{*}{Trick Prior} & $C_1$ & $\frac{C}{N}\frac{1}{H + V}$ \\
 & $C_2$ & $\frac{C}{N}\frac{1}{H*V}$ \\
\hline
\multirow{2}{*}{Truncated Normal} & $\sigma^2_{main}$ & $\frac{1}{H + V}$ \\
 & $\sigma^2_{interaction}$ & $\frac{1}{H*V}$ \\
\hline
\multirow{2}{*}{Marginalized Likelihood} & $\sigma^2_{main}$ & $\frac{1}{H + V}$ \\
 & $\sigma^2_{interaction}$ & $\frac{1}{H*V}$ \\
\hline
\end{tabular}
\caption{The values used for the hyperparameters for all three fitting methods. A rule of thumb is imposed which decreases as the size of the model increases and also shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.}
\label{tab:hyperparam}
\end{table}

Figure \ref{fig:fitting_plot} shows the posterior probability of each possible binary image with four pixels after fitting the model in the three ways detailed in this section. The black vertical lines show the true probabilities of each image based on the parameters used to simulate the images while the red vertical lines show the empirical distribution. From these posterior predictive checks, it is evident that the marginalized likelihood method shows the closest fit to the actual data. However, this model requires a marginalization step, which is infeasible for a model of any real size.

```{r fitting}
source("helpers/functions.R")
load("data/sample_images.Rdata")

#data and params ------------------------
H <- 4
V <- 4

#marginalized likelihood
load("data/fitted_models_trunc_marginal_thin.Rdata")
marginal_bad <- models_bad
marginal_good <- models_good

#load trick prior
load("data/fitted_models_jing_5.8.Rdata")
trick_bad <- models_bad
trick_good <- models_good

#truncated normal
load("data/fitted_models_trunc_thin.Rdata")
trunc_bad <- models_bad
trunc_good <- models_good

#rm unneccesary data
rm(models_bad)
rm(models_good)

load("data/params_theta.Rdata")

params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))

#computing actual distributions ---------------------
flat_images_good$visibles %>%
  data.frame() %>% 
  rename(v1 = X1, v2 = X2, v3 = X3, v4 = X4) %>%
  group_by(v1, v2, v3, v4) %>%
  summarise(prob = n()/nrow(flat_images_good$visibles)) -> distn_emp

distn_good <- visible_distn(params = params_good)

reshape_sample_distn <- function(model) {
  sample_distn <- model$distn
  dim(sample_distn) <- c(dim(sample_distn)[1]*dim(sample_distn)[2], dim(sample_distn)[3])
  sample_distn %>% data.frame() -> sample_distn
  names(sample_distn) <- names(distn_good)

@@

marginal_sample_good <- reshape_sample_distn(marginal_good)
trick_sample_good <- reshape_sample_distn(trick_good)
trunc_sample_good <- reshape_sample_distn(trunc_good)

marginal_sample_good %>% rename(marginal = prob) %>%
  left_join(trick_sample_good %>% ungroup() %>% select(-image_id) %>% rename(trick = prob)) %>%
  left_join(trunc_sample_good %>% ungroup() %>% select(-image_id) %>% rename(trunc = prob)) %>%
  ungroup() %>%
  mutate(image_id = paste0("image_", image_id)) %>% 
  gather(method, prob, trick, marginal, trunc) %>%
  spread(image_id, prob) -> all_statistics

all_statistics %>%
  gather(statistic, value, -iter, -method, -starts_with("v")) %>%
  filter(grepl("image", statistic) & !is.na(value)) %>%
  separate(statistic, into = c("junk", "statistic")) %>%
  mutate(statistic = factor(statistic, labels = paste("Image", 1:length(unique(statistic))))) %>%
  left_join(distn_emp %>% rename(prob_emp = prob) %>% right_join(distn_good %>% select(-image_id))) %>%
  ggplot() +
  geom_density(aes(value, y=..scaled.., colour = method, fill = method), alpha = .2) +
  geom_vline(aes(xintercept = prob)) +
  geom_vline(aes(xintercept = prob_emp), colour = "red") +
  facet_wrap(~statistic, scales = "free_x") + 
  ylab("Scaled Posterior Density") + xlab("Probability of Image") +
  scale_colour_discrete("Method", labels = c("Marginalized likelihood", "Trick prior", "Truncated Normal prior")) +
  scale_fill_discrete("Method", labels = c("Marginalized likelihood", "Trick prior", "Truncated Normal prior")) +
  theme(legend.position = "bottom") -> p.models
```

```{r fitting_plot, fig.cap="\\label{fig:fitting_plot}Posterior probability of each possible 4-pixel image using three Bayesian fitting techniques, trick prior, truncated Normal prior, and marginalized likelihood. The black vertical lines show the true probabilities of each image based on the parameters used to simulate the images while the red vertical lines show the empirical distribution. The marginalized likelihood method has the best fit for the data, however is also the most computationally intensive and least feasible with a real dataset.", fig.height=6}
p.models
```

@@

**TODO: Add proof that likelihood goes to infinity.**

@@

**TODO: Add comparison to contrastive divergence**
